\MakeUppercase{Layered $H$-Partitions of $k$-Planar Graphs} \thanks{This work was partly funded by NSERC and MRI.}

### 2019-05-16 k-planar.tex
Cubic bound seems to be working

The following claim is not used until later in the paper. However, it's proof is a helpful exercise in understanding the properties of a PS-Tree.

\begin{clm}\clmlabel{strong-ancestor-boundary}
  Let $K$ be a PS-tree and
  let $y\in V(K)$ be a non-root node with parent $p$ such that $V(F_y\cap Y_a)\neq\emptyset$ for some $K$-ancestor $a\neq p$ of $p$. Then $V(F_p\cap Y_a)\neq\emptyset$.
\end{clm}

\begin{proof}
  By \propref{children}, $F_y$ is a wedge in the Mercedes graph $M_p$, so $V(F_y)$ can be partitioned into  $P_1'=V(F_y\cap Y_p)$ and $P_{2,3}'=V(F_y\cap F_p)$.  Let $P_1,\ldots,P_g$ denote the partition of $V(F_p)$ given by \propref{ancestor-boundary}.

  We claim that \propref{tripod-feet} ensures that $P_{2,3}'\subset P_i\cup P_j$ for some $i,j\in \{1,\ldots,g\}$.  If $g=2$, this claim is immediate, so assume $g=3$. Then, $F_y\cap \overline{Y}_p$ is the path formed by two legs $Q_2$ and $Q_3$ of $\overline{Y}_p$ and one edge $v_2v_3$ of $\tau_p$.  
  By \propref{tripod-feet} and the assumption that $q=3$, $Q_2$ has a foot $v_2'\in P_i$ and $Q_3$ has a foot $v_3'\in P_j$, for some $i\neq j$.  The third foot $v_1'$ of $\overline{Y}_p$ is therefore in $P_\ell$ for the unique $\ell \in\{1,2,3\}\setminus \{i,j\}$.  It follows that $V(P_\ell\cap F_y)=\emptyset$.

  Setting $P_2'=V(P_i\cup F_y)$ and $P_3'=V(P_j\cup F_y)$ yields a partition $\{P_1',P_2',P_3'\}$ of $V(F_y)$ such that $P_1'\subset V(Y_{p})$,   
  $P_2'\subset V(Y_{a_2})$, and $P_3'\subset V(Y_{a_3})$  for some $K$-ancestors $a_2$ and $a_3$ of $a$.  Therefore $a_2$ and $a_3$ are the only $K$-ancestors $a\neq p$ of $p$, such that $V(F_y\cap Y_a)\neq\emptyset$.
  Now, each foot $v_2',v_3'\in V(F_p)$.  Furthermore, $v_2'\in P_i\subseteq V(Y_{a_2})$ and $v_3'\in P_j\subseteq V(Y_{a_3})$. Therefore, $V(F_p\cap Y_{a_2})\supseteq\{v_2'\}\neq\emptyset$ and $V(F_p\cap Y_{a_3})\supseteq\{v_3'\}\neq\emptyset$, as claimed.
\end{proof}

Contained in the proof of \clmref{strong-ancestor-boundary} is the following addendum to \propref{ancestor-boundary}:

\begin{enumerate}[(PR1')]\setcounter{enumi}{1}
  \item \proplabel{ancestor-boundary-ii} Furthermore, if $g=3$, then  $P_i=V(F_x\cap Y_p)$ where $p$ is the parent of $x$ in $K$, for one $i\in\{1,2,3\}$.
\end{enumerate}

@@

  \begin{figure}

    \begin{center}
      \begin{tabular}{cc}
        \includegraphics{figs/A-1} &
        \includegraphics{figs/A-3} \\ (a) & (b)
      \end{tabular}
    \end{center}
    \caption{(a)~the graph $A$ and (b)~the walks $W_{vw}$ and $W_{ww'}$ that contain the walk $w=v_0,v_1\ldots,v_{\ell-1},v_\ell=w^*$.}
    \figlabel{A}
  \end{figure}

@@

  To begin, let $d_0=d$, and $d_1=\min\{i: V(F_d\cap Y_i)\neq\emptyset \}$.  Observe that $F_{d}=F_{d_0}$ is contained in the interior of $F_{d_1}$, i.e., $N_{d_0}\subseteq N_{d_1}-F_{d_1}$.   We claim that 
  \[  
     |\{\delta\in\{d_1,\ldots,d_0-1\}:  \mbox{there exists a $\delta$-walk} \}| \le \binom{k+1}{2} \enspace .
  \]
  To prove this claim, consider the graph $I$ with vertex set $V(I)=\{d_1,\ldots,d_0\}$ and $ij\in E(I)$ if and only $i=j$ or $|F_i\cap Y_j|\neq\emptyset$.  Any $\delta$-walk $v_0,\ldots,v_\ell$ with $\delta\in V(I)$ corresponds to a walk $\rho_0,\ldots,\rho_{\ell}$ in $I$, where $\rho_i=\max\{j : v_i\in N_{j}-F_j\}$.

  The graph $I$ has the following properties:
  \begin{enumerate}
    \item For each $i\in V(I)$, $i$ has at most 2 neighbours among $0,\ldots,i-1$.  If $i$ has two such neighbours, then one of them is $i-1$.

    \item For each $ij\in E(I)$ with $j<i$, and each $i'\in\{j+2,\ldots,i\}$, the vertex $i'$ has at most 2 neighbours and these are contained in $\{i'-1,j\}$.
  \end{enumerate}
  These two properties imply that $I$ has a very simple structure.   We call $j\in\{d_1,\ldots,d_0\}$ a \emph{left endpoint} if there exists an edge $ij\in E(I)$ with $j < i - 1$.  Let $j_1>j_1>\cdots> j_p$ denote the sorted sequence of left endpoints and, for convenience, let $j_{0}=d$.  Then 
  \[
      E(I) \subseteq \{ (i,i+1): i\in\{d_1,\ldots,d_0-1\} \cup 
         \bigcup_{t=1}^p \{ ij_t : i\in\{ j_t,\ldots,j_{t-1} \}\} \enspace . 
  \]
  (See, \figref{i-graph}.)  Observe that shortest path from in $i$ from $d_0$ to $j_t$ has length at least $t$.  Let $X_t=\{j_t,\ldots,j_{t+1}\}$ and observe that the shortest path from $d_0$ to some value $i\in X_t$ that does not contain any $i' <i$ goes through $j_{t}$.  Therefore, $X_t$ contains at most $k+2-t$ values of $\delta$ such that there exists a $\delta$-walk.  Therefore, 
  \[  
     |\{\delta\in\{d_1,\ldots,d_0-1\}:  \mbox{there exists a $\delta$-walk} \}| \le \sum_{t=0}^{k+1} (k+2-t) = \sum_{t=1}^{k+2} t = \binom{k+2}{2} \enspace .
  \]

  \begin{figure}
    \begin{center}
      \includegraphics{figs/i-graph}
    \end{center}
    \caption{The graph $I$ used in the proof of \clmref{bag-size}}
    \figlabel{i-graph}
  \end{figure}

>>>>

% The following claim is not used until later in the paper. However, it's proof is a helpful exercise in understanding the properties of a PS-Tree.
% 
% \begin{clm}\clmlabel{strong-ancestor-boundary}
%   Let $K$ be a PS-tree and
%   let $y\in V(K)$ be a non-root node with parent $p$ such that $V(F_y\cap Y_a)\neq\emptyset$ for some $K$-ancestor $a\neq p$ of $p$. Then $V(F_p\cap Y_a)\neq\emptyset$.
% \end{clm}
% 
% \begin{proof}
%   By \propref{children}, $F_y$ is a wedge in the Mercedes graph $M_p$, so $V(F_y)$ can be partitioned into  $P_1'=V(F_y\cap Y_p)$ and $P_{2,3}'=V(F_y\cap F_p)$.  Let $P_1,\ldots,P_g$ denote the partition of $V(F_p)$ given by \propref{ancestor-boundary}.  
% 
%   We claim that \propref{tripod-feet} ensures that $P_{2,3}'\subset P_i\cup P_j$ for some $i,j\in \{1,\ldots,g\}$.  If $g=2$, this claim is immediate, so assume $g=3$. Then, $F_y\cap \overline{Y}_p$ is the path formed by two legs $Q_2$ and $Q_3$ of $\overline{Y}_p$ and one edge $v_2v_3$ of $\tau_p$.  
%   By \propref{tripod-feet} and the assumption that $q=3$, $Q_2$ has a foot $v_2'\in P_i$ and $Q_3$ has a foot $v_3'\in P_j$, for some $i\neq j$.  The third foot $v_1'$ of $\overline{Y}_p$ is therefore in $P_\ell$ for the unique $\ell \in\{1,2,3\}\setminus \{i,j\}$.  It follows that $V(P_\ell\cap F_y)=\emptyset$.
% 
%   Setting $P_2'=V(P_i\cup F_y)$ and $P_3'=V(P_j\cup F_y)$ yields a partition $\{P_1',P_2',P_3'\}$ of $V(F_y)$ such that $P_1'\subset V(Y_{p})$,   
%   $P_2'\subset V(Y_{a_2})$, and $P_3'\subset V(Y_{a_3})$  for some $K$-ancestors $a_2$ and $a_3$ of $a$.  Therefore $a_2$ and $a_3$ are the only $K$-ancestors $a\neq p$ of $p$, such that $V(F_y\cap Y_a)\neq\emptyset$.
%   Now, each foot $v_2',v_3'\in V(F_p)$.  Furthermore, $v_2'\in P_i\subseteq V(Y_{a_2})$ and $v_3'\in P_j\subseteq V(Y_{a_3})$. Therefore, $V(F_p\cap Y_{a_2})\supseteq\{v_2'\}\neq\emptyset$ and $V(F_p\cap Y_{a_3})\supseteq\{v_3'\}\neq\emptyset$, as claimed.
% \end{proof}
% 
% Contained in the proof of \clmref{strong-ancestor-boundary} is the following addendum to \propref{ancestor-boundary}:
% 
% \begin{enumerate}[(PR1')]\setcounter{enumi}{1}
%   \item \proplabel{ancestor-boundary-ii} Furthermore, if $g=3$, then  $P_i=V(F_x\cap Y_p)$ where $p$ is the parent of $x$ in $K$, for one $i\in\{1,2,3\}$.
% \end{enumerate}

@@

  \begin{figure}
    \begin{center}
      \begin{tabular}{cc}
        \includegraphics{figs/A-1} &
        \includegraphics{figs/A-3} \\ (a) & (b)
      \end{tabular}
    \end{center}
    \caption{(a)~the graph $A$ and (b)~the walks $W_{vw}$ and $W_{ww'}$ that contain the walk $w=v_0,v_1\ldots,v_{\ell-1},v_\ell=w^*$.}
    \figlabel{A}
  \end{figure}

@@

  To begin, let $d_0=d$, and $d_1=\min\{i: V(F_d\cap Y_i)\neq\emptyset \}$.  Observe that $F_{d}=F_{d_0}$ is contained in the interior of $F_{d_1}$, i.e., $N_{d_0}\subseteq N_{d_1}-F_{d_1}$.   Refer to \figref{d0d1}.  We claim that 
  \[  
     |\{\delta\in\{d_1,\ldots,d_0-1\}:  \mbox{there exists a $\delta$-walk} \}| \le \binom{k+1}{2} \enspace .
  \]
  \begin{figure}
    \begin{center}
      \includegraphics{figs/d0d1-5}
    \end{center}
    \caption{$F_{d_1},\ldots,F_{d_0}$ and the resulting graph $I$.}
    \figlabel{d0d1}
  \end{figure}
  To prove this claim, consider the graph $I$ with vertex set $V(I)=\{d_1,\ldots,d_0\}$ and $ij\in E(I)$ if and only $i=j$ or $|F_i\cap Y_j|\neq\emptyset$.  Any $\delta$-walk $v_0,\ldots,v_\ell$ with $\delta\in V(I)$ corresponds to a walk $\rho_0,\ldots,\rho_{\ell}$ in $I$, where $\rho_i=\max\{j\in\{0,\ldots,d\} : v_i\in N_{j}-F_j\}$.

  The graph $I$ has the following properties:
  \begin{enumerate}
    \item For each $i\in V(I)$, $i$ has at most 2 neighbours among $d_1,\ldots,i-1$.  If $i$ has two such neighbours, then one of them is $i-1$.

    \item For each $ij\in E(I)$ with $j<i$, and each $i'\in\{j+2,\ldots,i\}$, the vertex $i'$ has at most 2 neighbours and these are contained in $\{i'-1,j\}$.
  \end{enumerate}
  These two properties imply that $I$ has a very simple structure.   We call $j\in\{d_1,\ldots,d_0\}$ a \emph{left endpoint} if there exists an edge $ij\in E(I)$ with $j < i - 1$.  Let $j_1>j_2>\cdots> j_p$ denote the sorted sequence of left endpoints and, for convenience, let $j_{0}=d$.  Then 
  \[
      E(I) \subseteq \{ (i,i+1): i\in\{d_1,\ldots,d_0-1\} \cup 
         \bigcup_{t=1}^p \{ ij_t : i\in\{ j_t,\ldots,j_{t-1} \}\} \enspace . 
  \]
  (See the lower half of \figref{d0d1}.)  Observe that shortest path from in $i$ from $d_0$ to $j_t$ has length at least $t$.  Let $X_t=\{j_t,\ldots,j_{t+1}\}$ and observe that the shortest path from $d_0$ to some value $i\in X_t$ that does not contain any $i' <i$ goes through $j_{t}$.  Therefore, $X_t$ contains at most $k+2-t$ values of $\delta$ such that there exists a $\delta$-walk.  Therefore, 
  \[  
     |\{\delta\in\{d_1,\ldots,d_0-1\}:  \mbox{there exists a $\delta$-walk} \}| \le \sum_{t=0}^{k+1} (k+2-t) = \sum_{t=1}^{k+2} t = \binom{k+2}{2} \enspace .
  \]

  % \begin{figure}
  %   \begin{center}
  %     \includegraphics{figs/i-graph}
  %   \end{center}
  %   \caption{The graph $I$ used in the proof of \clmref{bag-size}}
  %   \figlabel{i-graph}
  % \end{figure}
